---
output: github_document
---

# rco - The R Code Optimizer

[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build status](https://travis-ci.org/jcrodriguez1989/rco.svg?branch=master)](https://travis-ci.org/jcrodriguez1989/rco)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/jcrodriguez1989/rco?branch=master&svg=true)](https://ci.appveyor.com/project/jcrodriguez1989/rco)
[![Coverage status](https://codecov.io/gh/jcrodriguez1989/rco/branch/master/graph/badge.svg)](https://codecov.io/github/jcrodriguez1989/rco?branch=master)

Make your R code run faster!
`rco` analyzes your code and applies different optimization strategies that return an R code that runs faster.

The `rco` project, from its start to version 0.1.008, was made possible by a [Google Summer of Code 2019 project](https://summerofcode.withgoogle.com/projects/#5337917017292800).

Thanks to the kind mentorship of [Dr. Yihui Xie](https://yihui.name/en/) and [Dr. Nicolás Wolovick](https://cs.famaf.unc.edu.ar/~nicolasw/).

```{r echo=FALSE}
library("rco")
```

## Installation

`rco` is currently only available as a GitHub package.

To install it run the following from an R console:

```{r eval = FALSE}
if (!require("remotes")) {
  install.packages("remotes")
}
remotes::install_github("jcrodriguez1989/rco", dependencies = TRUE)
```

## Usage

`rco` can be used in two ways:

* Using the RStudio Addins

  1. `Optimize active file`: Optimizes the file currently open in RStudio. It will apply the optimizers present in `all_optimizers`.
  
  2. `Optimize selection`: Optimizes the code currently highlited in the RStudio Source Pane. It will apply the optimizers present in `all_optimizers`.

* Using the R functions

  1. Optimize some `.R` code files
  
  ```{r eval = FALSE}
  optimize_files(c("file_to_optimize_1.R", "file_to_optimize_2.R"))
  ```
  
  2. Optimize some code in a character vector
  
  ```{r eval = FALSE}
  code <- paste(
    "code_to_optimize <- 8 ^ 8 * 1918",
    "cto <- code_to_optimize * 2",
    sep = "\n"
  )
  optimize_text(code)
  ```

## Example

Suppose we have the following code:

```{r}
code <- paste(
  "# I want to know my age in seconds!",
  "years_old <- 29",
  "days_old <- 365 * years_old # leap years don't exist",
  "hours_old <- 24 * days_old",
  "seconds_old <- 60 * 60 * hours_old",
  "",
  "if (seconds_old > 10e6) {",
  '  print("Whoa! More than a million seconds old, what a wise man!")',
  "} else {",
  '  print("Meh!")',
  "}",
  sep = "\n"
)
```

We can automatically optimize it by doing:

```{r}
opt_code <- optimize_text(code, iterations = 1)
```

```{r echo = FALSE}
cat(opt_code)
```

After one optimization pass we can see that it has only propagated the `years_old` variable. Another pass:

```{r}
opt_code <- optimize_text(opt_code, iterations = 1)
```

```{r echo = FALSE}
cat(opt_code)
```

Now, it has folded the `days_old` variable, and then propagated it. Another pass:

```{r}
opt_code <- optimize_text(opt_code, iterations = 1)
```

```{r echo = FALSE}
cat(opt_code)
```

It has folded the `hours_old` variable, and then propagated it. Another pass:

```{r}
opt_code <- optimize_text(opt_code, iterations = 1)
```

```{r echo = FALSE}
cat(opt_code)
```

It has folded the `seconds_old` variable, and then propagated it into the `if` condition. Another pass:

```{r}
opt_code <- optimize_text(opt_code, iterations = 1)
```

```{r echo = FALSE}
cat(opt_code)
```

Now, it has folded the `if` condition, and as it was `TRUE` it just kept its body, as testing the condition or the `else` clause were dead code.
So, `optimize_text` function has automatically detected constant variables, constant foldable operations, and dead code. And returned an optimized R code.

## Guidelines for contributing

`rco` is an open source package, and the contributions to the development of the library are more than welcome. Please see our [CONTRIBUTING.md](.github/CONTRIBUTING.md) file and ["Contributing an Optimizer"](https://jcrodriguez1989.github.io/rco/articles/contributing-an-optimizer.html) article for detailed guidelines of how to contribute.

## Code of Conduct

Please note that the 'rco' project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md).

By contributing to this project, you agree to abide by its terms.
