---
output: rmarkdown::html_vignette
title: Constant Folding
vignette: >
  %\VignetteIndexEntry{Constant Folding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
library("rco")
library("microbenchmark")
library("ggplot2")
autoplot.microbenchmark <- function(obj) {
    levels(obj$expr) <- paste0("Expr_", seq_along(levels(obj$expr)))
    microbenchmark:::autoplot.microbenchmark(obj)
}
speed_up <- function(obj) {
    levels(obj$expr) <- paste0("Expr_", seq_along(levels(obj$expr)))
    obj <- as.data.frame(obj)
    summaries <- do.call(rbind, by(obj$time, obj$expr, summary))
    res <- c()
    for (i in seq_len(nrow(summaries)-1)+1) {
        res <- rbind(res, summaries[1,] / summaries[i,])
    }
    rownames(res) <- levels(obj$expr)[-1]
    return(res)
}
```

# Constant Folding

## Background

[Source](https://en.wikipedia.org/wiki/Constant_folding)

Constant folding is the process of recognizing and evaluating constant expressions rather than computing them at runtime. Consider the statement: 

`i <- 320 * 200 * 32`

Most compilers would not actually generate two multiply instructions. Instead, they identify constructs such as these and substitute the computed values (in this case, `2048000`).

## Example

A simple example would be to have to convert the unit of many temporary samples from hours to miliseconds `miliseconds <- 1000 * 60 * 60 * hours`.

```{r}
code <- paste(
  "hours_vector <- runif(1000, 0, 24)",
  "ms_vector <- numeric(1000)",
  "# of course it would be much efficient to do vectorized operations xP",
  "for (i in seq_along(hours_vector)) {",
  "  ms_vector[i] <- 1000 * 60 * 60 * hours_vector[i]",
  "}",
  sep = "\n")
cat(code)
```

Then, the automatically optimized code would be:

```{r}
opt_code <- opt_constant_folding(list(code))
cat(opt_code$codes[[1]])
```

And if we measure the execution time of each one, and the speed-up:

```{r}
bmark_res <- microbenchmark(
  {
    eval(parse(text = code))
  },{
    eval(parse(text = opt_code))
  }
)
autoplot(bmark_res)
speed_up(bmark_res)
```

## Implementation

Actually, `opt_constant_folding` will fold expressions that are conformed solely by operators and constants which tokens are (`utils::getParseData`):

```{r echo=FALSE}
rco:::ops
rco:::constants
```

## Floating-point precision

When constant folding, we can have floating-point precision issues, for instance, consider the following code:

```{r}
x <- 1/(2+1)
y <- 1/(2+1)
z <- 1/(2+1)
x + y + z == 1
```

If we constant fold it, we would have:

```{r}
opt_code <- opt_constant_folding(list(paste(
  "x <- 1/(2+1)",
  "y <- 1/(2+1)",
  "z <- 1/(2+1)",
  "x + y + z == 1",
  sep = "\n")))$codes[[1]]
cat(opt_code)
```

However, this code is not equivalent due to precision:

```{r}
x <- 0.333333333333333
y <- 0.333333333333333
z <- 0.333333333333333
x + y + z == 1
```

In this case, we can use the parameter `fold_floats`. If set to FALSE, then the optimizer will fold every expression except those which fold loses precision:

```{r}
opt_code <- opt_constant_folding(list(paste(
  "x <- 1/(2+1)",
  "y <- 1/(2+1)",
  "z <- 1/(2+1)",
  "x + y + z == 1",
  sep = "\n")), fold_floats = FALSE)$codes[[1]]
cat(opt_code)
```

Consider this nice example where a sub-expression folding causes precision loss, but as it is not important in overall, then it can be folded:

```{r}
opt_code <- opt_constant_folding(list(paste(
  "x <- 1/(2+1)",     # will not fold it because we lose precision
  "y <- 1/(2+1) > 3", # however, folded or not, it is not > 3, so folds it
  sep = "\n")), fold_floats = FALSE)$codes[[1]]
cat(opt_code)
```

## To-Do

* Implement intelligent constant folding?
  
  For example: fold `0 * x` to `0`
  
  However, this could change code behavior, in the second case, if `x` does not exist then the code would not throw an error, meanwhile, in the first case, it would.

* Reorder variables with associative operators?
  
  The R parser has left associativity, so `x + 10 + 200` is `((x + 10) + 200)`. So this is not being folded to `x + 210`.
  
  If we consider operators with associativity, we could replace `x + 10 + 200` to  `10 + 200 + x`, and then fold it to `210 + x`

* Try constant folding if a function takes constants?
  
  For example: If we have `paste("aa", "bb")`, we know that this can be folded to `"aa bb"`. The optimizer could try function calls that have all constants as input.
  
  However, this could break the optimizer, for instance, if we try to optimize the following code (when evaluating `foo(3)`):
  
  ```{r eval=FALSE}
  foo <- function(x) {
    while(TRUE) {
      x <- x+1
    }
  }
  foo(3)
  ```
