---
output: rmarkdown::html_vignette
title: Efficient Column Extraction
vignette: >
  %\VignetteIndexEntry{Dead Code Elimination}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Efficient Column Extraction
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Column extraction refers to the act of pulling out the values of a particular column from a given data frame/list/matrix. The below microbenchmark plot clearly shows the vast difference in execution time between `.subset2()` and other methods of column extraction. Therefore, it would make sense to replace other column extraction methods with `.subset2()` since it is a function call while all others are function calls.

```{r Microbenchmark plot, echo=FALSE, message=FALSE, warning=FALSE}
library("microbenchmark")
library("ggplot2")
autoplot(microbenchmark(
  mtcars[,11],
  mtcars$carb,
  mtcars[[c(11)]],
  mtcars[[11]],
  .subset2(mtcars, 11)
))
```

For example, consider the following code:

```{r raw_code, eval=FALSE}
  mtcars[,11]
  mtcars$carb
  replace_text <- c(1,2,3,4,5)
  .subset2(mtcars, 11)
```

There is nothing wrong with the above code, but as noticed earlier it can be made efficient, if we replace the given column extraction methods with system calls(`.subset2()`). That is, it can be made more efficient in the following way:

```{r opt_code, eval=FALSE}
  .subset2(mtcars, 11)
  .subset2(mtcars, 11)
  replace_text <- c(1,2,3,4,5)
  .subset2(mtcars, 11)

```

## Implementation

The `opt_column_extractor` function of the `rco` achieves this by, parsing the given code snippet, detecting the usage of column extraction methods/techniques and replacing it with the most appropriate one.