---
title: "Efficient Value Extraction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Efficient Column Extraction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Value extraction refers to the act of pulling out a particular value represented by a row and column given data frame/list/matrix. The below microbenchmark plot clearly shows the vast difference in execution time between **.subset2()[]** and other methods of value extraction. Therefore, it would make sense to replace other value extraction methods with **.subset2()[]** since it is a function call while all others are function calls.

```{r Microbenchmark plot, echo=FALSE, message=FALSE, warning=FALSE}
library("microbenchmark")
library("ggplot2")
autoplot(microbenchmark(
  mtcars[32, 11],
  mtcars$carb[32],
  mtcars[[c(11, 32)]],
  mtcars[[11]][32],
  .subset2(mtcars, 11)[32]
))
```

For example, consider the following code:

```{r raw_code, eval=FALSE}
  mtcars[32, 11]
  mtcars$carb[32]
  replace_text <- c(1,2,3,4,5)
  .subset2(mtcars, 11)[32]
```

There is nothing wrong with the above code, but as noticed earlier it can be made efficient, if we replace the given row extraction methods with system calls(.subset2()[]). That is, it can be made more efficient in the following way:

```{r opt_code, eval=FALSE}
  .subset2(mtcars, 11)[32]
  .subset2(mtcars, 11)[32]
  replace_text <- c(1,2,3,4,5)
  .subset2(mtcars, 11)[32]
```

## Implementation

The *opt_value_extractor* function of the *rco* achieves this by, parsing the given code snippet, detecting the usage of value extraction methods/techniques and replacing it with the most appropriate one, by singling out the row_number and column_number in all cases individually.